Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-02 paper/paper.tex
very minor tweaks to results section 6

It remains unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Specifically, because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have genetic defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and more IFN is induced by viral stocks that contain defective virions with internal deletions in the polymerase genes~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

@@

\subsection{Viral features associated with infection outcome in single cells}

@@

Viral features also associated with IFN induction, with cells infected by full wildtype virions turning IFN+ less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
Specific viral defects were especially associated with IFN induction (\FIG{mutations}C).
Absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched.
Only the absence of NS remained significant at a false discovery rate of 10\% (\FIG{mutations}C)---although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent, since some are more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
In any case, \FIG{genotypes} and \FIG{mutations}C motivate compelling hypotheses about specific viral defects that contribute to IFN induction, and below we pursue these hypotheses using experimental follow-up.

@@

The viral defect most strongly associated with IFN induction in single cells was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cell lines to generate influenza viruses lacking a specific viral gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza virus's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.

@@

Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a large deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
We therefore created a virus carrying this deletion that we were able to propagate in cells constitutively expressing the PB1 protein.

We next tested the rate of IFN induction by each viral stock using the reporter cell lines.
As shown in \FIG{validation}, five of the eight mutant viruses induced IFN more frequently than wildtype.
The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the six point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wildtype.
These results show that the viral defects in single IFN+ cells often cause increased IFN production.

It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of particular viral defects among the IFN+ cells in the single-cell data.
However, these comparisons are confounded by the fact that the measurements in \FIG{validation} are made using \emph{bulk} stocks of each viral variant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
For instance, our single-cell experiment used a ``wildtype'' viral stock, but the vast majority of IFN+ cells (31 of 40) still failed to express wildtype copies of all viral genes (\FIGSUPP[genotypes]{genotypes_by_ifn}). 
Therefore, although the increased IFN induction by most viral variant stocks in \FIG{validation} does indicate that these variants are causally associated with IFN, without single-cell virus-sequencing it is impossible to know how many of the IFN+ cells induced by each viral stock are infected by virions that also have other defects.
For instance, most of the 0.67\% of cells infected with the ``wildtype'' viral stock that turned IFN+ in \FIG{validation}A are probably infected by virions that have some type of genetic defect.

>>>>

It remains unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
Such self-imposed limitation of an IFN response is consistent with the biological impact of IFN as, despite the beneficial antiviral effects of IFN signaling, these effects can come at a significant cost due to increased immunopathology~\citep{iwasaki2014innate}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Specifically, because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have genetic defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and more IFN is induced by viral stocks that contain defective virions with internal deletions in the polymerase genes~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

@@

\subsection{Viral defects associated with infection outcome in single cells}

@@

Viral defects also associated with IFN induction, with cells infected by full wildtype virions turning IFN+ less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
Specific viral defects were especially associated with IFN induction (\FIG{mutations}C).
Absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched.
Only the absence of NS remained significant at a false discovery rate of 10\% (\FIG{mutations}C)---although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent, since some are more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
In any case, \FIG{genotypes} and \FIG{mutations}C motivate compelling hypotheses about specific viral defects that contribute to IFN induction, and below we pursue these hypotheses using experimental follow-up.

@@

The viral defect most strongly associated with IFN induction in single cells was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza virus's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.

@@

Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.

We tested the rate of IFN induction by each viral stock using the reporter cells.
\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wildtype.
The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wildtype.
These results show that the viral defects in single IFN+ cells often cause increased IFN production.

It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of particular viral defects among the IFN+ cells in the single-cell data.
However, these comparisons are confounded by the fact that the measurements in \FIG{validation} use \emph{bulk} stocks of each viral mutant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
For instance, our single-cell experiment used a ``wildtype'' viral stock, but most IFN+ cells (31 of 40) still had viral defects (\FIGSUPP[genotypes]{genotypes_by_ifn}). 
So although increased IFN induction by the mutant viral stocks does indicate the mutations are causally associated with IFN, it is impossible to know how many of the IFN+ cells induced by each stock are infected by virions that also have other defects.
The rate of IFN induction by any given viral mutant can therefore only be accurately quantified by single-cell virus-sequencing.
